---
title: "EEB313 project code"
author: "Jessica Bullock"
date: "2024-11-18"
output: pdf_document
---
This project aims to determine the importance of lineage and ecology in grouping Apistogramma dwarf cichlids by phenotype. 
This code was written to analyse the combined data from 'Apisto_ecological_data.csv', 'Apisto_character_data.csv', and 'Apisto_lineage_data.csv'.

#Load required packages
```{r, message=F, warning=F}
require(tidyverse)
library(FactoMineR) # to perform MCA
library(factoextra) # to plot MCA stuff
library(qgraph) # to create the qgraph, showing correlation between nodes in a matrix
library(ape) # to make NJ tree (was not successful, made.. a stick)
library(phangorn) # to make UPGMA tree
```

Load individual data files, merge by species.
Change '?' to 'NA' as no explanation was given by the author.
Change 'occaisional' to 'yes', as data is based on presence/absence. (and I can't spell).

```{r}
eco_data<- read.csv("Apisto_ecological_data.csv")
character_data<- read.csv("Apisto_character_data.csv")
lineage_data<- read.csv("Apisto_lineage_data.csv")

book_data <- merge(eco_data, character_data, by="species")
total_data<- merge(book_data, lineage_data, by="species")


total_data[total_data == '?']<-'NA'
total_data[total_data == 'occaisional']<-'yes'
```

Need to filter out species with missing data, then rename rows as species names, and remove species column.
Remove columns with only one level remaining; they will not help inform analysis.
Confirm structure of data before analysis.

```{r}
total_data %>% 
  filter(if_all(everything(), ~!str_detect(., "NA")))%>%
  mutate_if(is.numeric,as.character)->data

rownames(data) <- data$species
data$species <- NULL


final_data <- data[, sapply(data, function(x) length(unique(x)) > 1)]


final_data<-as.data.frame(final_data)
str(final_data)
```

Run Multiple Correspondence analysis, save biplot of individuals as jpg.
'repel = TRUE' helps prevent overlapping labels

```{r}
results_mca<-MCA(final_data)

summary(results_mca)

factor_map_plot <- fviz_mca_ind(results_mca, repel = TRUE, 
            ggtheme = theme_minimal())

jpeg("mca_factor_map.jpg", width = 800, height = 600)  
print(factor_map_plot) 
dev.off()
```

Visualize the percentages of inertia explained by each MCA dimension in a scree plot, reduce y axis, save as jpg

```{r}
scree_plot <-fviz_screeplot(results_mca, addlabels = TRUE, ylim = c(0, 25))

jpeg("scree_plot.jpg", width = 800, height = 600)  
print(scree_plot) 
dev.off()
```
Contribution of each variable category (in %) to each dimension

```{r}
var <- get_mca_var(results_mca)

var$contrib
```

#Create plot of variable corelation to dimensions 1 and 2, save jpg

```{r}
var_plot<- fviz_mca_var(results_mca, choice = "mca.cor", 
            repel = TRUE, labels= TRUE, labelsize = 6,
            ggtheme = theme_minimal())

jpeg("var_plot.jpg", width = 1200, height = 900)  
print(var_plot) 
dev.off()
```

Dimension description - identify the most correlated variables with each dimension, along with their r^2.

```{r}
res_desc <- dimdesc(results_mca, axes = c(1,2))

res_desc[[1]]

res_desc[[2]]
```

#Create Bi-Plots of species, group and colour by water types and then by lineage. Save as jpgs.
Ellipses denote 95% confidence around the means of both dimensions 1 and 2 for the selected category.
Colours chosen from accessible pallettes at http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette.

By Blackwater
```{r}
blackwater_plot <- fviz_mca_ind(results_mca, 
             label = "none", 
             habillage = "blackwater", 
             palette = c("#888888","#D55E00"),
             addEllipses = TRUE, ellipse.type = "confidence", 
             ggtheme = theme_minimal()) +
  ggtitle("Species grouped by variable: Blackwater")

jpeg("blackwater_plot.jpg", width = 1200, height = 900)  
print(blackwater_plot) 
dev.off()
```

By Whitewater
```{r}
whitewater_plot <- fviz_mca_ind(results_mca, 
             label = "none", 
             habillage = "whitewater", 
             palette = c("#888888", "#D55E00"),
             addEllipses = TRUE, ellipse.type = "confidence",
             ggtheme = theme_minimal()) +
   ggtitle("Species grouped by variable: Whitewater")

jpeg("whitewater_plot.jpg", width = 1200, height = 900)  
print(whitewater_plot) 
dev.off()
```

By Clearwater
```{r}
clearwater_plot <- fviz_mca_ind(results_mca, 
             label = "none", # hide individual labels
             habillage = "clearwater", # DO BY WATER TYPE color by level
             palette = c("#888888", "#D55E00"),
             addEllipses = TRUE, ellipse.type = "confidence",
             ggtheme = theme_minimal()) +
   ggtitle("Species grouped by variable: Clearwater")

jpeg("clearwater_plot.jpg", width = 1200, height = 900)  
print(clearwater_plot) 
dev.off()
```

By Lineage
```{r}
lineage_plot <- fviz_mca_ind(results_mca, 
             label = "none", # hide individual labels
             habillage = "lineage", # DO BY WATER TYPE color by level
             palette = c("#009E73","#56B4E9","#F0E442", "#D55E00"),
             addEllipses = TRUE, ellipse.type = "confidence",
             ggtheme = theme_minimal()) +
   ggtitle("Species grouped by variable: Lineage")

jpeg("lineage_plot.jpg", width = 1200, height = 900)  
print(lineage_plot) 
dev.off()
```


#Extract Coordinates/Loadings for Apisto species ('individuals') with get_mca_ind()

```{r}
ind <- get_mca_ind(results_mca)
ind

head(ind$coord)
```

#Create a new dataframe of Apisto species and their coordinates/loadings from Dimensions 1 and 2 of the MCA, to provide continuous data for analysis.
```{r}
ind_loadings <- ind$coord[, 1:2]
loadings <- data.frame(species = rownames(final_data), Dimension_1 = ind_loadings[, 1], Dimension_2 = ind_loadings[, 2])
head(loadings)
```

# Create standard distance matrix (Euclidean) from Dimension 1 loadings
```{r}
dm <- dist(loadings[,2])
```

#Create QGraph to show strength of relationships between species based on the distance matrix from the dimension 1 loadings. 
Thicker lines denote stronger correlation.
Create list of species names to label nodes.
Create list of lineage values, convert to numeric format, assign list of colours to lineages.Colour species by Lineage.
Plot qgraph and save as jpeg, add legend and title.
```{r}
species_names <- rownames(final_data)

lineage_values <- final_data$lineage
lineage_values<-as.numeric(lineage_values)
lineage_colors <-  c("#009E73","#56B4E9","#F0E442", "#D55E00")
label_colors <- lineage_colors[lineage_values]


jpeg("qgraph_mca.jpg", width = 1400, height = 900)
qgraph(dist(loadings[,2]), layout='spring', vsize=5, 
       labels = species_names, color = label_colors, label.cex = 1.5)
par(mar = c(1, 1, 1, 1))

legend("topright",                       # Position of the legend
       legend = c("Lineage 1", "Lineage 2", "Lineage 3", "Lineage 4"), 
       fill = lineage_colors,             # Color corresponding to each lineage
       title = "Lineage",                 # Title for the legend
       bty = "n",                         #
       cex = 0.8)     
title( main = "Strength of correlation between Apistogramma species")
dev.off()
```

#MARCH OF THE ENTS
Try making a tree from the distance matrix using the neighbour-joining method.
Create and plot UPGMA tree using the distance matrix from the loadings, save as jpg.
Add title, tip labels of the species names, colour names by lineage, and legend
```{r}
eco_NJ<- NJ(dm)
plot(eco_NJ)
#YIKES, lets try something else

eco_UPGMA <- upgma(dm)

jpeg("eco_tree_lineages.jpg", width = 1600, height = 900)
par(mar = c(4, 4, 4, 14))
plot(eco_UPGMA, main="Combined data tree with Tougard et al lineages, UPGMA method")
tiplabels(text=rownames(final_data), frame = "n", cex=0.8, col= label_colors, offset = 0.025)
unique_lineages <- unique(lineage_values)
unique_colors <- lineage_colors[unique_lineages]
legend("topleft", 
       legend = unique_lineages,               
       fill = unique_colors,                   
       cex = 0.8,                              
       title = "Lineage",                     
       bty = "n",                              
       inset = c(0.02, 0.02))
dev.off()
```

Citations
```{r}
citation()
citation("tidyverse")
citation("FactoMineR")
citation("factoextra")
citation("qgraph")
citation("ape")
citation("phangorn")
```
Thanks for reading, best fishes on your next analysis.
